Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2017-11-17 paper/paper.tex
First draft of paper with mostly written sections.

%% packages
\usepackage{graphicx,psfrag}
\usepackage{mathrsfs,amsmath,amsfonts,amssymb, bm}
\usepackage{multirow}
\usepackage{comment,hyperref}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algpseudocode} % in place of algorithmic. Conflicts with revtex4 unless [H] option passed
\usepackage{setspace} % for pleasent spacing in algorithm
\usepackage{times}

@@

Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters. The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as corrections for dynamical tides, that are accurate enough to avoid significant biases in the measured parameters. However, current implementations take at least 10 minutes to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations. We construct a frequency-domain surrogate model that requires less than 0.2~s to evaluate for a waveform with a starting frequency of 10~Hz (0.05~s for a starting frequency of 30~Hz). We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter. We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with approximately 500 waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods. For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.

%Finally, I will show how this same method can be used to optimally choose parameters for numerical relativity simulations to correct the the EOB surrogate during the last several orbits and merger.

@@

%% Template for figures
%% \begin{figure}[t]
%%   \centering 
%%     \includegraphics[width=0.49\textwidth]{../fig/ }
%%     \caption{ }
%%  \label{fig: }
%% \end{figure}

%% Template for tables
%% \begin{table}[t]
%%   \centering    
%%   \caption{ }
%%   \begin{tabular}{ccc}        
%%     \hline
%%     col$1$ & col$2$ & col$3$ \\
%%     \hline
%%     &  &  \\
%%     \hline
%%   \end{tabular}
%%  \label{tab: }
%% \end{table}

\red{[[Change arrow vectors to bold vectors. ${\bm x}^i$ (bold) for example $i$. $x_j$ (not bold) for coordinate/parameter/feature $j$. 
$x_j^i$ (not bold) for feature $j$ of example $i$.]]}

@@

The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO) [[cite]] now making routine detections of binary black hole (BBH) systems [[cite]], as well as Virgo [[cite]], KAGRA [[cite]], and LIGO-India [[cite]]. Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]], and the nondetection of BNS events in aLIGO's first observing run (O1) is still consistent with the upper end of this range to 90\% confidence [[O1UpperLimits]]. However, continued nondetection in runs now underway will decrease the upper limit of this range.

The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters. These parameters, measured with Bayesian parameter estimation techniques, require highly accurate waveform models to avoid systematic errors. For example, post-Newtonian (PN) waveform models, which are truncated at the currently known 3.5PN order, lead to uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz [[cite]]. This is approximately the same size as the contribution to the phase from tidal effects, and as a result, the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]]. This also biases the recovered NS equation of state and radius [[cite]], as well as the masses and spins [[cite]].

The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms. Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad. Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]]. Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]]. Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable. Significant work has been done to optimize existing implementations of EOB for BBH systems [[EtienneMcWilliams]]. However, it is not clear that these optimizations can be readily adapted for the BNS model discussed here with dynamical tides. We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.

@@

@@

@@

\subsection{Approximation of higher-order tidal effects}

@@

As discussed above in Sec.~II, the current implementation of the EOB model takes at least 10~minutes to evaluate. In this section we will describe in detail the method for training and evaluating our surrogate model. The first decision in building a waveform surrogate is choosing between building a surrogate of the time-domain waveform $h(t;\bx)$ or the frequency-domain waveform $\tilde h(f;\bx)$. A time-domain surrogate has the advantage that significantly less processing of the original waveform is needed, resulting in a surrogate that typically more accurately reproduces the original waveform model. A frequency-domain surrogate, on the other hand, requires windowing of the Fourier transformed waveform to avoid Gibbs oscillations. For the beginning of the waveform, one can simply begin at a lower frequency, window the waveform, then truncate the Fourier-transformed waveform above some desired starting frequency. At the end of the waveform, the BNS amplitude drops rapidly over $\sim 1$~GW cycle, resulting in Gibbs oscillations at the high-frequency end. One could window the end of the waveform over several cycles, however, this would remove available information about tidal interactions. As we will show in Sec.~[[]], these high-frequency Gibbs oscillations in the amplitude of the Fourier transformed waveform are a genuine feature of BNS waveforms unlike BBH waveforms which typically taper over a larger number of GW cycles during their ringdown. These oscillations make it difficult to construct a surrogate. However, we will show that smoothing these oscillations in the Fourier transformed waveform with a moving average filter leads to a surrogate that is sufficiently faithful to the original time-domain model. Another important advantage is that a surrogate of $\tilde h(f;\bx)$ bypasses the need to evaluate the Fourier transform at each waveform evaluation for parameter estimation. And, as we shall see, a frequency-domain surrogate can be trivially attached to analytic models that are accurate at low-frequencies, such as TaylorF2, making analysis for third-generation detectors which are sensitive at lower frequencies easier. Finally, methods for speeding up the likelihood evaluation in parameter estimation such as reduced order quadrature typically require a fast frequency-domain model for calculating the appropriate weights online [[cite]]. For these reasons we will construct a surrogate of $\tilde h(f;\bx)$.

%The problem of constructing a surrogate model for a computer simulation that takes parameters 
%$\vec x$ and outputs a vector of data $\vec y$ is the following: construct approximate functions $f_j(x)$
%for each quantity $y_j$ using a training set of precomputed examples $\{\vec x^i, \vec y^i\}$. For the problem here, 
%our parameters are the intrinsic waveform parameters $\vec x$ and the data is the waveform frequency series $\tilde h(f_j)$.

@@

\item Construct separate orthonormal, reduced bases for the amplitude $A(f;\bx)$ and phase $\Phi(f;\bx)$ of the training set waveforms. Then, use the empirical interpolation method (EIM) to generate separate sets of frequency-dependent empirical interpolating functions for the amplitude $B^A_j(f)$ and phase $B^\Phi_j(f)$.

\item For each empirical interpolating function, interpolate the amplitude $A(F^A_j;\bx)$ or phase $\Phi(F^\Phi_j;\bx)$ at the associated frequency nodes $F^A_j$ or $F^\Phi_j$, respectively, as a function of the waveform parameters $\bx$ using GPR.

@@

\subsection{Initial training set design}

@@

\subsection{Conditioning and decomposing the training set waveforms}

--Messy details of preconditioning lalsimulation waveforms to make them smooth functions of parameters ${\bm x}$.
(Rescale with total mass to convert to geometric units, time shift to align at max amplitude or specific frequency, 
taper beginning and end, pad end with zeros, Fourier transform, truncation, resampling, zeroing initial phase, ...)

--Show plot (?) to validate that waveform is properly conditioned by plotting amplitude/phase as function of waveform parameters
at specific frequencies and show that they're smooth functions.

\begin{equation}
\tilde h(f, x) = A(f, x) e^{i\Phi(f, x)}
\end{equation}

Because the waveform $h$ is an oscillatory function of $f$ and ${\bm x}$ that is difficult to interpolate, we instead construct 
surrogates for the amplitude $A(f, x)$ and phase $\Phi(f, x)$. These are smooth functions. However, for aligned-spin waveforms
these functions are already approximately known from the 3.5PN TaylorF2 waveform 
$\tilde h_{\rm F2}(f, x) = A_{\rm F2}(f, x) e^{i\Phi_{\rm F2} (f, x)}$, where
\begin{equation}
A_{\rm F2} = \sqrt{\frac{5\pi\eta}{6}} x^{-7/4},
\end{equation}
and
\begin{align}
\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
\end{align}
Where $x=(\pi f)^{2/3}$ is the standard PN parameter. We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~[[]] of Ref.~[[]], the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order in Eqs.~[[]] of Ref.~[[]] and Eqs.~[[]] of Ref.~[[]], and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order in Eqs.~[[]] of Ref.~[[]]. We therefore only have to construct surrogates for the difference between EOB and TaylorF2:
\begin{align}
\Delta\ln(A) &= \ln(A/A_{\rm F2})\\
\Delta\Phi & = \Phi - \Phi_{\rm F2}
\end{align}
These quantities are shown in Fig.~\ref{fig:dh} for the training set wavefroms. We note that for small differences 
between the EOB and TaylorF2 waveform, $\Delta\ln A$ is approximately the fractional difference $A/A_{\rm F2} - 1$
between the EOB and TaylorF2 waveform. We use the log-ratio instead of the ratio, because it guarantees that any
error in interpolating this quantity doesn't result in a negative amplitude. Also, because waveforms tend to deviate from 
each other near the end of the inspiral exponentially, depending on each waveform's merger frequency the log captures
this behavior. [[Is this true for EOB? We'll find out.]]

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
\caption{Training set}
\label{fig:dh}
\end{figure}

We show the range of values of the phase $\Phi$ over the training set $\Phi_{\rm range} = \Phi_{\rm max} - \Phi_{\rm min}$ 
as well as the range of values of the phase residual $\Delta\Phi$ over the training set 
$\Delta\Phi_{\rm range} = \Delta\Phi_{\rm max}-\Delta\Phi_{\rm min}$ in Fig.~\ref{fig:phaserange}. 
The range in $\Phi$ can be larger than 1000~radians, meaning
that if we require 0.1~radian accuracy, we need a fractional error of $<10^{-4}$. The range of $\Delta\Phi$ is 1--2 orders 
of magnitude smaller, and as a result, the interpolation accuracy requirements are 1--2 orders of magnitude smaller as well.

\subsection{Compression of waveform data with reduced bases and empirical interpolation method}

For many surrogate problems a function $f({\bm x})$ is sought to interpolate a scalar quantity $y$. For the
problem here, the outputs are the quantities $\Delta\ln A$ and $\Delta\Phi$ sampled at the frequencies $f_j$. 
One could interpolate $\Delta\ln A$ and $\Delta\Phi$ as a function of ${\bm x}$ at each sample $f_j$, 
however, this would be prohibitively expensive. Instead we use the standard method of constructing an
orthonormal reduced basis to compress each training set waveform into a small set of coefficients, 
and then using the empirical interpolation method
to optimize the placement of a corresponding number of empirical nodes $F^A_j$ and $F^\Phi_j$. One then interpolates
$\Delta\ln A$ and $\Delta\Phi$ as a function of ${\bm x}$ at these nodes. The full waveform can then be reconstructed by 
multiplying the value at these nodes by the the corresponding empirical interpolating functions $B_j^A(f)$ and $B_j^\Phi(f)$. 
The methods used here are presented in Refs.[[cite]], and we use the exact implementation described in 
Ref.~[[LackeyBernuzziGalley2017]]. We briefly describe the details.

>>>>

%% packages
\usepackage{graphicx,psfrag}
\usepackage{mathrsfs,amsmath,amsfonts,amssymb, bm}
\usepackage{multirow}
\usepackage{comment,hyperref}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algpseudocode} % in place of algorithmic. Conflicts with revtex4 unless [H] option passed
\usepackage{setspace} % for pleasent spacing in algorithm
\usepackage{times}
\usepackage{ulem}

@@

Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters. The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as the effect of dynamical tides, and the formalism is accurate enough to avoid significant biases in the measured parameters. Unfortunately, current implementations take several minutes to a day to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations. We construct a frequency-domain surrogate of this model that requires less than 0.3~s to evaluate for a waveform with a starting frequency of 10~Hz (0.04~s for a starting frequency of 30~Hz). We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter. We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with $\sim 1000$ waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods.

For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.

@@

@@

\red{[[What is possible with BNS systems:]]}

The detection of the binary neutron star (BNS) system GW170817 [[cite]] by the Advanced LIGO [[cite]] and Advanced Virgo [[cite]] gravitational-wave detectors, has demonstrated that the sky position, distance, masses, spins, and tidal parameters can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant [[cite]], and understanding the structure of neutron stars (NSs) through tidal interactions [[cite]]. Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction. Furthermore, an estimate of the BNS merger rate of x--y events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval) [[GW170817]] indicates that, when second generation detectors reach design sensitivity, including KAGRA [[cite]] and possibly LIGO-India [[cite]], we could eventually observe a population of tens or hundreds of events after a couple of years of observation [[2010Rates?]]. This will allow us to estimate the NS mass distribution [[cite]], spin distribution [[cite]], and universal equation of state (EOS) that is common to all NSs [[DelPozzo, Lackey, ...]].

\red{[[Why we need more accurate waveform models to do these things:]]}

These results are obtained using a Bayesian analysis that compares the data to parameterized waveform templates. To avoid systematic errors highly accurate waveform templates are needed. In addition, the Bayesian parameter estimation alogrithms (Markov chain Monte Carlo and Nested Sampling [[parallel algorithms are becoming available]]) typically perform several million sequential waveform evaluations, requiring waveform evaluation times to be less than $\mathcal{O}(1s)$ in order to run in less than a month for each event.

%The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters. These parameters, measured with Bayesian parameter estimation techniques, require highly accurate waveform models to avoid systematic errors. For example, post-Newtonian (PN) waveform models, which are truncated at the currently known 3.5PN order, lead to uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz [[cite]]. This is approximately the same size as the contribution to the phase from tidal effects, and as a result, the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]]. This also biases the recovered NS equation of state and radius [[cite]], as well as the masses and spins [[cite]].

\red{[[The problems with the current waveform models/templates.]]}

To date, there are a limited number BNS waveform models that include spin and tides, and that are fast enough for standard parameter estimation.  
TaylorF2: (not so good point particle. okay tidal terms)
For example, post-Newtonian (PN) waveform models, which are truncated at the currently known 3.5PN order, lead to uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz [[cite]]. This is approximately the same size as the contribution to the phase from tidal effects, and as a result, the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]]. This also biases the recovered NS equation of state and radius [[cite]], as well as the masses and spins [[cite]].

\*\_NRTides: (better/good point particle. better tidal terms)
Initial parameter estimation results for the first BNS event, GW170817 using a variety of waveform templates (TaylorF2, SEOBNRv4\_ROM\_NRTides, and IMRPhenomD\_NRTides) showed that there are noticeable differences in the measured parameters between the different waveform templates [[cite]], and in particular the tidal parameter. [[NR\_Tides is a fit to NR waveforms at high frequency and has only been fit to a limited number of low spin, low tidal parameter NR simulations.]]

TEOBResumS:

\red{[[What is good about the new EOB model:]]}

The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms. Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad. Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]].

Another model

\red{[[Problem with EOB (speed). Attempts to speed it up with optimization. Alternative using surrogates.]]}

Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]]. Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable. Significant work has been done to optimize existing implementations of EOB for BBH systems [[EtienneMcWilliams]]. However, it is not clear that these optimizations can be readily adapted for the BNS model discussed here with dynamical tides. We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.

@@

\red{[[Overview of surrogates. Typical methods that have been used. Their problems. What you are doing that is new.]]}

%The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms. Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad. Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]]. Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]]. Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable. Significant work has been done to optimize existing implementations of EOB for BBH systems [[EtienneMcWilliams]]. However, it is not clear that these optimizations can be readily adapted for the BNS model discussed here with dynamical tides. We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.

%Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$. The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions. In previous work, this reduced basis has been constructed either with singular value decomposition (SVD) [[Michael's papers]] or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization [[All the other papers]].

@@

\red{[[Obligitory summary:]]}

@@

\subsection{Approximation of higher-order tidal parameters}

@@

%As discussed above in Sec.~II, the current implementation of the EOB model takes at least 10~minutes to evaluate. In this section we will describe in detail the method for training and evaluating our surrogate model. The first decision in building a waveform surrogate is choosing between building a surrogate of the time-domain waveform $h(t;\bx)$ or the frequency-domain waveform $\tilde h(f;\bx)$. A time-domain surrogate has the advantage that significantly less processing of the original waveform is needed, resulting in a surrogate that typically more accurately reproduces the original waveform model. A frequency-domain surrogate, on the other hand, requires windowing of the Fourier transformed waveform to avoid Gibbs oscillations. For the beginning of the waveform, one can simply begin at a lower frequency, window the waveform, then truncate the Fourier-transformed waveform above some desired starting frequency. At the end of the waveform, the BNS amplitude drops rapidly over $\sim 1$~GW cycle, resulting in Gibbs oscillations at the high-frequency end. One could window the end of the waveform over several cycles, however, this would remove available information about tidal interactions. As we will show in Sec.~[[]], these high-frequency Gibbs oscillations in the amplitude of the Fourier transformed waveform are a genuine feature of BNS waveforms unlike BBH waveforms which typically taper over a larger number of GW cycles during their ringdown. These oscillations make it difficult to construct a surrogate. However, we will show that smoothing these oscillations in the Fourier transformed waveform with a moving average filter leads to a surrogate that is sufficiently faithful to the original time-domain model. Another important advantage is that a surrogate of $\tilde h(f;\bx)$ bypasses the need to evaluate the Fourier transform at each waveform evaluation for parameter estimation. And, as we shall see, a frequency-domain surrogate can be trivially attached to analytic models that are accurate at low-frequencies, such as TaylorF2, making analysis for third-generation detectors which are sensitive at lower frequencies easier. Finally, methods for speeding up the likelihood evaluation in parameter estimation such as reduced order quadrature typically require a fast frequency-domain model for calculating the appropriate weights online [[cite]]. For these reasons we will construct a surrogate of $\tilde h(f;\bx)$.

\subsection{Decomposition of the waveform}

In this section we will describe the method for decomposing the waveform into smooth, slowly-varying functions and training a surrogate model for these functions. As discussed above, we work with the frequency domain waveform $\tilde h(f; \bx)$. This is favorable for data analysis which is usually done in the frequency domain. However, Fourier transforming a finite length, discretely sampled waveform, typically leads to a surrogate with more noise. We will show below, that sufficient filtering can solve most of these problems. Because the waveform $\tilde h(f; \bx)$ is an oscillatory function of $f$ and $\bx$, the waveform is usually decomposed into an amplitude and phase $\tilde h(f; \bx) = A(f; \bx) e^{i\Phi(f; \bx)}$, which are smoother, mostly monotonic functions of $f$ and $\bx$. The amplitude and phase of these waveforms are shown in Fig.~\ref{fig:h}.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{h_trainingset.png}\\
\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
\caption{Training set}
\label{fig:h}
\end{figure}

Although the amplitude and phase are relatively smooth functions, they still span a wide range of values. The phase, for example, spans about $10^4$~rad. Unfortunately, to avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad. This is a difficult requirement to achieve in the 5-dimensional parameter space here. However, for aligned-spin waveforms these functions are approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 $\tilde h_{\rm F2}(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2} (f; \bx)}$ [[cite]]. We can therefore evaluate the EOB waveform in terms of the residuals, $\Delta\ln(A)$ and $\Delta\Phi$, between EOB and TaylorF2
\begin{align}
\tilde h(f; \bx) = A_{\rm F2}(f; \bx) e^{i\Phi_{\rm F2}(f; \bx)} e^{\Delta\ln(A)(f; \bx) + i  \Delta\Phi(f; \bx)},
\label{eq:hdecomp}
\end{align}
where
\begin{align}
\Delta\ln(A)(f; \bx) &= \ln\left(\frac{ A(f; \bx) }{ A_{\rm F2}(f; \bx)}\right),\\
\Delta\Phi(f; \bx) & = \Phi(f; \bx) - \Phi_{\rm F2}(f; \bx).
\end{align}
These residuals are shown in Fig.~\ref{fig:h}. For the amplitude, we use the log-ratio instead of the ratio because it guarantees that the amplitude of the reconstructed waveform in Eq.~\eqref{eq:hdecomp} is positive even if interpolation error leads to a negative $\Delta\ln(A)(f; \bx)$. In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency of each waveform, the log-ratio captures this behavior.

To demonstrate the usefulness of this approach, in Fig.~\ref{fig:phaserange}, we show the range of values of the phase $\Phi$ over the training set $\Phi_{\rm range} = \Phi_{\rm max} - \Phi_{\rm min}$ as well as the range of values of the phase residual $\Delta\Phi$ over the training set $\Delta\Phi_{\rm range} = \Delta\Phi_{\rm max}-\Delta\Phi_{\rm min}$. The range in $\Phi$ can be larger than 1000~radians. The range of $\Delta\Phi$ is 1--2 orders of magnitude smaller, and as a result, the interpolation accuracy requirements are 1--2 orders of magnitude smaller as well.

The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
\begin{equation}
A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
\end{equation}
where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{lalsuite}. The phase has the schematic form
\begin{align}
\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
\end{align}
We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~[[]] of Ref.~[[]], the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order in Eqs.~[[]] of Ref.~[[]] and Eqs.~[[]] of Ref.~[[]], and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order in Eqs.~[[]] of Ref.~[[]]. These terms for the phase are exactly as used in the \texttt{lalsuite} waveform \texttt{TaylorF2} [[cite]].

\subsection{Preconditioning the training set waveforms}

We evaluate the EOB waveform with a starting frequency equivalent to 20Hz for a total mass of $M=2M_\odot$ ($Mfwi=0.000197$). In order to take a discrete Fourier transform, we window the start of the waveform with a Planck window [[cite]] in the interval $(Mfwi, Mfwf) = (0.000197, 0.00021)$ to avoid Gibbs oscillations. We then resample the waveform with a spacing $\Delta t/M = 5$, and pad the end of the waveform with zeros such that all waveforms in the training set have the exact same time samples.

After evaluating the discrete Fourier transform, evaluate the residuals $\Delta\ln(A)$ and $\Delta\Phi$ between the EOB and TaylorF2 waveforms. Because waveforms have free time and phase parameters, subtract a linear fit to $\Delta\Phi$ at the beginning of the waveform in the window $(Mf_{fit,i}, Mf_{fit,f}) = 0.00021(1, 1.05)$. The resulting $\Delta\Phi$ is zero and has zero slope at $Mf_{fit,i}$, guaranteeing that the surrogate smoothly matches to TaylorF2 at lower frequencies.

As a result of Gibbs oscillations in the Fourier transformed waveform that weren't completely removed by the Planck waveform as well as from the end of the waveform where the amplitude rapidly drops to zero amplitude after the peak amplitude over $\sim 1$ cycle, small amplitude, high frequency oscillations remain in the residuals $\Delta\ln(A)$ and $\Delta\Phi$. We remove these oscillations using a moving average filter centered on $Mf$ with width $[Mf(1-\Delta Mf), Mf(1+\Delta Mf)]$. We use a width of $\Delta Mf=0.1$ for the amplitude and $\Delta Mf=0.05$ for the phase.

Finally, we truncate the data to the interval $(Mf_{trunc,i}, Mf_{trunc,f}) = (0.00021, 0.07)$. We note that the gravitational wave frequency at the Schwarzschild ISCO is $Mf_{\rm ISCO} = 1/(6^{3/2}\pi) \approx 0.022$. Although second generation detectors are only marginally sensitive to signals above this frequency, we use the high-frequency cutoff of $Mf_{trunc,f}=0.07$ because some of the EOB waveforms still have enough amplitude at these high frequencies to affect the structure of the surrogate waveform if inverse Fourier transformed back into the time domain.

\red{[[Show plot (?) to validate that waveform is properly conditioned by plotting amplitude/phase as function of waveform parameters at specific frequencies and show that they're smooth functions?]]}

\subsection{Spline interpolation for frequency $f$}

We now seek to interpolate the conditioned and filtered residuals $\Delta\ln(A)(f; \bx)$ and $\Delta\Phi(f; \bx)$. We begin by interpolating as a function of $f$ for fixed $\bx$. The majority of previous papers on gravitational waveform surrogates have used an orthonormal basis of global functions $\hat e_i(f)$ for frequency (or time) interpolation [[cite]]. This type of surrogate is referred to as a reduced order model. For a generic function $g(f; \bx)$, this can be written as $g(f; \bx) \approx \sum_{i=1}^N c_i(\bx) \hat e_i(f)$. The coefficients $c_i(\bx)$ are then interpolated as functions of $\bx$ [[Puerrer]]. Alternatively, using the empirical interpolation method [[cite]], one can re-express these basis functions $\hat e_i(f)$ in terms of empirical interpolating functions $B_j(f)$ and the value of $g(f; \bx)$ at empirical nodes $F_j$ as $g(f; \bx) \approx \sum_{j=1}^N B_j(f) g(F_j; \bx)$. The location of these nodes $F_j$ can be optimized to minimize interpolation errors. This method has been used in Refs.~[[cite]].

For the problem here, we have found that global basis functions do not work well. The residuals are small and smooth at low frequencies and large and noisy at high frequencies. Errors in evaluating $g(F_j;\bx)$ for high frequency can propagate to large errors at low frequencies. Instead, we find that spline interpolation works significantly better. We use 20 frequency nodes $F_j$ log-spaced the interval $[f_{trunc,i}, f_{trunc,f}]$, and evaluate the residuals $\Delta\ln(A)(F_j; \bx)$ and $\Delta\Phi(F_j; \bx)$ at these nodes using GPR discussed below. We then choose third-order spline interpolation to interpolate between these points. The local, third-order polynomials, that are only connected by the requirement of smoothness [[check this]] do not propagate high-frequency errors down to low-frequency errors as significantly.

\subsection{Gaussian process regression for parameters $\bx$}

We use Gaussian process regression to fit the residuals $\Delta\ln(A)(F_j; \bx)$ and $\Delta\Phi(F_j; \bx)$ at the empirical nodes $F_j$ as a function of the waveform parameters. In GPR a function $f(\vec x)$ is described in terms of a mean $m(\vec x)$ and 
covariance $k(\vec x, \vec x')$ between points [[cite RassmussenWilliamsChapter2, IntroArxivPaper]].
\begin{equation}
f(\vec x) \approx \mathcal{GP}(m(\vec x), k(\vec x, \vec x')).
\end{equation}
The mean $m(x)$ can be some parameterized function which is the standard version of regression, 
while $k(\vec x, \vec x')$ is a kernel with tunable hyperparameters that describes the covariance between
the point $\vec x$ and the sampled point $\vec x'$. This covariance can be used to describe features of the function 
not captured by the parameterized mean as well as genuine noise in the data ${\bm y}$. 
For the problem here, we have already subtracted the 
TaylorF2 waveform from EOB waveform, so we set $m=0$ and model the remainder terms purely in terms 
of the covariance $k$.

\subsection{Iterative construction of training set}

@@

%\item Construct separate orthonormal, reduced bases for the amplitude $A(f;\bx)$ and phase $\Phi(f;\bx)$ of the training set waveforms. Then, use the empirical interpolation method (EIM) to generate separate sets of frequency-dependent empirical interpolating functions for the amplitude $B^A_j(f)$ and phase $B^\Phi_j(f)$.

\item Choose interpolation nodes.

%\item For each empirical interpolating function, interpolate the amplitude $A(F^A_j;\bx)$ or phase $\Phi(F^\Phi_j;\bx)$ at the associated frequency nodes $F^A_j$ or $F^\Phi_j$, respectively, as a function of the waveform parameters $\bx$ using GPR.

\item GPR

@@

@@
